\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-07-09 submission/manuscript.Rmd
prep for sarah and add interp paragraph

We explained the feature importances in non-linear models, SVM with radial basis kernel, decision tree, random forest and XGBoost, using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect of permuted OTUs or group of OTUs on testing AUROC. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure S3]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].

We used permutation importance to interpret linear models as well to highlight the differences between the two interpretation methods [Figure S3]. The rankings of OTUs that discriminate between having an SRN or not, changed between different interpretation methods. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings [Figure 3] and permutation importance methods [Figure S3]. However, in general L2-regularized models had different OTUs picked out as important in the two methods of interpretation due to collinearity in microbial communities.

@@

As the complexity of a ML model and the number of tuned hyperparameter settings increased [Table S1], its training times increased as well [Figure 5]. Linear models trained faster than non-linear models. L1 and L2 SVM with linear kernel and L2 logistic regression had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, a decision tree, SVM with radial basis function kernel, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

@@

In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. We set-up standards for developing and evaluating ML models for microbiome data. First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the differences between cross-validation and test performance is low, this suggets the models are not overfit and that they will perform similar with similar data. In all seven models, the differences in cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models will be able to test similar new data. Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop generalizable ML models and to report reliable performance metrics for a ML model. Depending on the data-split, there is the chance of being overoptimistic about the predictive performance of a model. In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.Third,  we used AUROC metric instead of accuracy in our study to evaluate predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced. We also performed a full grid search for hyperparameter settings when building a ML model. It is not enough to use the default hyperparameter settings when using previously developed ML packages in programming languages such as R, Python and Matlab. In the example of L1-regularized SVM with linear kernel (Figure S1), the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

Our results showed that we should choose to use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden. In terms of predictive performance, random forest model had testing AUROC values statistically significantly larger than others. However the second best model was L2-regularized logistic regression following random forest with a median AUROC difference of only 0.015. In terms of interpretation, random forest was a more complex ML model and it could only be explained using methods such as permutation importance. On the other hand, L2-regularized logistic regression was easy to interpret (i.e. regression coefficients of the trained model). If the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers of a disease, then easily interpretable models would be better suited. It is also important to consider the computational burden of developing ML models. Random forest model was slow to train 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes. Another criteria when choosing a ML model is generalizability. The generalization performance of ML models depends on size. The more complex the model, the more data it will need. The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.

Microbiome studies use ML models with a classification task to learn the training data to assign labels, healthy or not, to new data (e.g. new samples), but also to learn which features are important enough to discriminate between labels. This is a difficult task because we need to correctly interpret our ML models. In this study we used two methods to interpret our models. Absolute feature weights of trained models and permutation importance. When interpreting linear models with the differences we observed between the two interpretation methods is possibly due to collinearity in the dataset. Collinearity is that one OTU is highly correlated with other OTUs which means that the regression coefficients/weights are not unique and that they are influenced by each other. If they are influenced and not unique then it becomes harder to interpret the model. You may want to use different models for prediction and one for interpretation/inference, in one you remove highly correlated OTUs to interpret and in other keep them all to predict.

>>>>

We explained the feature importances in non-linear models, SVM with radial basis kernel, decision tree, random forest and XGBoost, using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect of permuted OTUs or group of OTUs on testing AUROC. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure S3]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].

We used permutation importance to interpret linear models as well, to highlight the differences between the two interpretation methods [Figure S3]. The rankings of OTUs that discriminate between having an SRN or not, changed between different interpretation methods. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings [Figure 3] and permutation importance methods [Figure S3] but with different ranks. The difference was more pronounced in L2-regularized models as they had mostly different OTUs picked out as important due to collinearity in microbial communities.

@@

As the complexity of a ML model [Table S1] and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5]. Linear models trained faster than non-linear models. L1 and L2 SVM with linear kernel and L2 logistic regression had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, a decision tree, SVM with radial basis function kernel, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

@@

In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. We set-up standards for developing and evaluating ML models for microbiome data. First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the difference between cross-validation and test performance is low, this suggets the models are not overfit and that they will perform similar with similar data. In all seven models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data. Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics for a ML model. Depending on the data-split, there is the chance of being overoptimistic about the predictive performance of a model. In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split. Third, we used the AUROC metric instead of accuracy in our study to evaluate the predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced. We also performed a full grid search for hyperparameter settings when building a ML model. It is not enough to use the default hyperparameter settings when using previously developed ML packages in programming languages such as R, Python and Matlab. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden. Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not, to new data, but also to learn which features are important to discriminate between labels. If the goal of a study is not just to predict a health outcome but also to learn the ecology behind a disease and to identify microbial biomarkers of a disease, then the ML model has to be interpretable. In terms of predictive performance, random forest model had testing AUROC values statistically significantly larger than others. However the second best model was L2-regularized logistic regression following random forest with a median AUROC difference of only 0.015. In terms of interpretability, random forest was a complex ML model and it could only be explained using methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret (i.e. ranking absolute regression coefficients of the trained model).

Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models. In this study we used two methods to interpret our linear models; absolute feature weights of trained models and permutation importance. We observed differences in outcome between the two interpretation methods due to collinearity in the dataset. Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes regression coefficients to not be unique. The regression coefficients of correlated OTUs are influenced by one another which makes it difficult to interpret a ML model. There are ways to make it easier to interpret models.; (1) use different models for prediction and interpretation in which highly correlated OTUs are removed, (2) get the highly ranked OTUs and check what othe OTUs they are correlated with. The interpretation however should not be a final answer but should be used to generate hypothesis that are tested with followup experiments.

There are other criteria when choosing ML models such as the computational burden of developing it, sample size and balance of the dataset. In terms of computation burden, random forest model was slow to train 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes. The generalization performance of ML models depends on sample size. The more complex the model, the more data it will need. The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.

### 2019-07-09 submission/manuscript.tex
prep for sarah and add interp paragraph

We used permutation importance to interpret linear models as well to
highlight the differences between the two interpretation methods
{[}Figure S3{]}. The rankings of OTUs that discriminate between having
an SRN or not, changed between different interpretation methods.
L1-regularized SVM with linear kernel picked out some of the same OTUs
(OTU00822, OTU01239, OTU00609) as important in feature rankings
{[}Figure 3{]} and permutation importance methods {[}Figure S3{]}.
However, in general L2-regularized models had different OTUs picked out
as important in the two methods of interpretation due to collinearity in
microbial communities.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We set-up standards
for developing and evaluating ML models for microbiome data. First, we
used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the differences between
cross-validation and test performance is low, this suggets the models
are not overfit and that they will perform similar with similar data. In
all seven models, the differences in cross-validation and testing AUROC
values did not exceed 0.021 which suggests that these models will be
able to test similar new data. Second, we performed the initial
80\%-20\% random datasplit 100 times in our ML pipeline. The
randomization of the initial data-split to create a held-out test set is
a crucial step in the ML pipeline to develop generalizable ML models and
to report reliable performance metrics for a ML model. Depending on the
data-split, there is the chance of being overoptimistic about the
predictive performance of a model. In our study, we showed that there
was variability in AUROC values between different random data-splits in
each of the models we tested. Our results showed that the testing AUROC
values varied 0.23 on average, depending on the data-split.Third, we
used AUROC metric instead of accuracy in our study to evaluate
predictive performance of the ML models. AUROC is always random at the
value 0.5 and is a robust metric when a dataset is imbalanced. We also
performed a full grid search for hyperparameter settings when building a
ML model. It is not enough to use the default hyperparameter settings
when using previously developed ML packages in programming languages
such as R, Python and Matlab. In the example of L1-regularized SVM with
linear kernel (Figure S1), the model showed large variability between
different regularization coefficients (C) and was susceptible to
performing poorly if the wrong regularization coefficient was assigned
to the model by default.

Our results showed that we should choose to use ML models based on the
goal of the study and our expectations of predictive performance,
interpretibility and computational burden. In terms of predictive
performance, random forest model had testing AUROC values statistically
significantly larger than others. However the second best model was
L2-regularized logistic regression following random forest with a median
AUROC difference of only 0.015. In terms of interpretation, random
forest was a more complex ML model and it could only be explained using
methods such as permutation importance. On the other hand,
L2-regularized logistic regression was easy to interpret
(i.e.~regression coefficients of the trained model). If the goal of a
study is to learn the ecology behind a disease and to identify microbial
biomarkers of a disease, then easily interpretable models would be
better suited. It is also important to consider the computational burden
of developing ML models. Random forest model was slow to train 83.2
hours whereas L2-regularized logistic regression trained in 12 minutes.
Another criteria when choosing a ML model is generalizability. The
generalization performance of ML models depends on size. The more
complex the model, the more data it will need. The dataset we used for
our study had 490 samples, however microbiome studies that have smaller
sample sizes would benefit from using less complex models such as
L2-regularized logistic regression.

Microbiome studies use ML models with a classification task to learn the
training data to assign labels, healthy or not, to new data (e.g.~new
samples), but also to learn which features are important enough to
discriminate between labels. This is a difficult task because we need to
correctly interpret our ML models. In this study we used two methods to
interpret our models. Absolute feature weights of trained models and
permutation importance. When interpreting linear models with the
differences we observed between the two interpretation methods is
possibly due to collinearity in the dataset. Collinearity is that one
OTU is highly correlated with other OTUs which means that the regression
coefficients/weights are not unique and that they are influenced by each
other. If they are influenced and not unique then it becomes harder to
interpret the model. You may want to use different models for prediction
and one for interpretation/inference, in one you remove highly
correlated OTUs to interpret and in other keep them all to predict.

>>>>

We used permutation importance to interpret linear models as well to
highlight the differences between the two interpretation methods
{[}Figure S3{]}. The rankings of OTUs that discriminate between having
an SRN or not, changed between different interpretation methods.
L1-regularized SVM with linear kernel picked out some of the same OTUs
(OTU00822, OTU01239, OTU00609) as important in feature rankings
{[}Figure 3{]} and permutation importance methods {[}Figure S3{]} but
with different ranks. The difference was more pronounced in
L2-regularized models as they had mostly different OTUs picked out as
due to collinearity in microbial communities.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We set-up standards
for developing and evaluating ML models for microbiome data. First, we
used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the differences between
cross-validation and test performance is low, this suggets the models
are not overfit and that they will perform similar with similar data. In
all seven models, the differences in cross-validation and testing AUROC
values did not exceed 0.021 which suggests that these models are
generalizable and can be used to test similar new data. Second, we
performed the initial 80\%-20\% random datasplit 100 times in our ML
pipeline. The randomization of the initial data-split to create a
held-out test set is a crucial step in the ML pipeline to develop robust
ML models and to report reliable performance metrics for a ML model.
Depending on the data-split, there is the chance of being overoptimistic
about the predictive performance of a model. In our study, we showed
that there was variability in AUROC values between different random
data-splits in each of the models we tested. Our results showed that the
testing AUROC values varied 0.23 on average, depending on the
data-split. Third, we used AUROC metric instead of accuracy in our study
to evaluate predictive performance of the ML models. AUROC is always
random at the value 0.5 and is a robust metric when a dataset is
imbalanced. We also performed a full grid search for hyperparameter
settings when building a ML model. It is not enough to use the default
hyperparameter settings when using previously developed ML packages in
programming languages such as R, Python and Matlab. In the example of
L1-regularized SVM with linear kernel (Figure S1), the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

In this study, we benchmarked seven ML models with different
classification algorithms to show that we should use ML models based on
the goal of the study and our expectations of predictive performance,
interpretibility and computational burden. Microbiome studies use ML
models with the classification task to learn the training data to assign
labels, such as healthy or not, to new data (e.g.~new samples), but also
to learn which features are important enough to discriminate between
labels. If the goal of a study is not just to predict a health outcome
but also to learn the ecology behind a disease and to identify microbial
biomarkers of a disease, then interpretibility of a ML model is crucial.
In terms of predictive performance, random forest model had testing
AUROC values statistically significantly larger than others. However the
second best model was L2-regularized logistic regression following
random forest with a median AUROC difference of only 0.015. In terms of
interpretability, random forest was a more complex ML model and it could
only be explained using methods such as permutation importance. On the
other hand, L2-regularized logistic regression was easier to interpret
(i.e.~ranking absolute regression coefficients of the trained model).

Even with interpretable models, we need to be careful with how we treat
the information we get from the models. In this study we used two
methods to interpret our linear models; absolute feature weights of
trained models and permutation importance. We observed differences in
outcome between the two interpretation methods due to collinearity in
the dataset. Collinearity in a microbial dataset is when one OTU is
highly correlated with another OTU which causes regression coefficients
to not be unique. The regression coefficients of correlated OTUs are
influenced by one another which makes it difficult to interpret a ML
model. There are ways to make it easier to interpret models.; (1) use
different models for prediction and interpretation in which highly
correlated OTUs are removed, (2) get the highly ranked OTUs and check
what othe OTUs they are correlated with. The interpretation however
should not be a final answer but should be used to generate hypothesis
that are tested with followup experiments.

There are other criteria when choosing ML models such as the
computational burden of developing it, sample size and balance of the
dataset. In terms of computation burden, random forest model was slow to
train 83.2 hours whereas L2-regularized logistic regression trained in
12 minutes. The generalization performance of ML models depends on
sample size. The more complex the model, the more data it will need. The
dataset we used for our study had 490 samples, however microbiome
studies that have smaller sample sizes would benefit from using less
complex models such as L2-regularized logistic regression.
